---
title: "Project 1 Report"
author: "Anthony Chiado, Benjamin Nikolai, Grace Zhang, & Reanna Panagides"
date: "2023-06-30"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning=FALSE, message=FALSE)
```

```{r, echo=FALSE}
#loading the data
library(tidyverse)
diamonds<-read.csv("diamonds4.csv", header=TRUE)
```

## 1. Key Findings

Blue Nile, Inc is a diamond and engagement ring business that claims to help prospective diamond buyers choose the highest-quality diamond at the best value. According to the Blue Nile Website, diamond buyers should use the 4Cs of diamonds when considering a purchase: cut, color, clarity, and carat. Regarding diamond cut, Blue Nile claims that diamonds with the “Astor Ideal" cut are of highest value. Additionally, diamonds with less color hue are generally more expensive. Blue Nile also claims that diamonds with fewest and smallest imperfections have the highest clarity grades and the highest prices. And finally, they state that diamonds with higher carat weights have the highest prices.

Blue Nile makes many strong claims about how these 4Cs impact the stone value but are they true? A study conducted in 2023 investigates if these claims are supported or unsupported by diamond pricing data. This study explores the relationships between the 4Cs and price from 1214 diamonds that are for sale on the Blue Nile online store.

This study found that for each additional carat size increasing, diamond price was multiplied by $6.99. In support of Blue Nile’s claims, these findings reveal that as the carat weight of a diamond increases, so does its price. Similarly, the data supports the claim that diamonds with less color due are generally higher in price. Conversely to Blue Nile’s claims, higher-quality diamond cut does not necessarily increase the price. Using average prices alone, “Good” diamond cut has the highest prices compared to the highest-quality cut, “Astor Ideal”. Additionally, the data shows that diamonds with the highest clarity grade are not necessarily higher than diamonds that have lower grades. Clarity grades VS1, VVS2, and VVS1, which are moderately graded diamonds, are priced the highest.

Overall, this study supports claims that diamonds with higher carat weight and more colorless hue have the highest prices. The data also shows that higher quality diamond cut and clarity are not typically priced the highest. This study suggests that when making a diamond purchase, diamond carat and color appear to have the greatest contribution to a diamond's price. 

## 2. Analysis

### 2.1 Overview
The Blue Nile website heavily highlights the importance every variable has on the price of a diamond, with similar emphasis on each being the most significant factor. Following this, our analysis on this data first focused on creating visualizations to represent and determine the effect that each variable has on price independently. This would allow us to gain an understanding of how strongly each variable affects price, and would give us knowledge relationships to tackle further claims the website makes. Next, if we noticed a relationship between “C’s”, we performed an analysis to determine if any noticeable patterns occur between diamond characteristics.

**Main Blue Nile Claims Investigated During this Study Include:**

1) The higher the diamond carat, the higher the price. 
2) The better the cut, the higher the price.
3) The more colorless the diamond, the higher the price.
4) The less imperfections (higher clarity grade), the higher the price.

**Other Relationships Between Diamond Characteristics Investigated During this Study Include:**

1) Relationship between Cut, Carat, and Price
2) Relationship between Color, Carat, and Price
3) Relationship between Color and Clarity

### 2.2 Carat

#### 2.2.1 Variable Description and Claim

First, we investigated the relationship between diamond carat and price. Blue Nile claims that diamond carat size has the biggest effect on a diamond's value, where diamonds with higher carat weights are more expensive. In this study, diamond carat refers to a diamond’s weight. One diamond carat is equal to 200 mg or 0.2 grams. In our sample, the median carat weight was 0.52, with a minimum carat weight of 0.23 and a maximum carat weight of 7.09.

```{r, echo=FALSE, results='hide'}
#summary(diamonds$carat)
```

#### 2.2.2 Findings

According to the scatter plot analysis in **Figure 2.2.1** there exists a non-linear relationship between the two data points where as carat weight increases by one, the diamond price is multiplied by $6.99. If one were to fit a logistic regression line to the data, they might also notice the variance of price the data also increases as the carat increases as well. Because of this, for further visualizations throughout this report, we perform a log transformation on the price variable. It is also worth to note that the majority of the diamond observations exist with lower carat weights, it is possible that if we could use more observations with higher carats, a clear pattern would be visible in the higher price ranges. Although the relationship between carat and price is non-linear, generally speaking, as carat increases so does price. In general, the data support's Blue Nile's claims about diamond carat and price. 

```{r, echo=FALSE}
#Scatter plot of Carat vs. Price
diamonds %>%
  ggplot(aes(x=carat, y=price))+
  geom_point(alpha=0.2)+
  theme(plot.caption = element_text(hjust = 0.5, face = "bold"))+
  labs(x="Diamond Carat", y="Price ($)", caption="Figure 2.2.1 Scatter Plot of Diamond Carat Against Price")
```

### 2.3 Cut

#### 2.3.1 Variable Description and Claim

Second, we investigated the relationship between diamond cut and price. Blue Nile claims that the "Astor Ideal" cut is the highest valued cut while the "Good" cut is the lowest valued cut. In this study, diamond cut refers to how well-proportioned the dimensions of the diamond are and how its surfaces are positioned to create sparkle and brilliance. Our data set is broken into 4 different sets of cut, in order of increasing value they are: “Good”, “Very Good”, “Ideal”, and “Astor Ideal”. In our data set, the most popular diamond cut is the "Ideal" cut (60.9%), followed by the "Very Good" cut (31.5%), then "Good" cut (6.0%), and finally the "Astor Ideal" cut (1.6%).

```{r, echo=FALSE}
#summary(diamonds$cut)
#round(prop.table(summary(diamonds$cut)), 3)
```


#### 2.3.2 Findings

According to **Figure 2.3.1** for our price, Astor Ideal cuts have a range of values with the highest median of all cuts, while Ideal stretches to the lowest to include the values with the lowest median. Both diamonds with "Good" and Very Good" cuts have higher median prices than diamonds with "Ideal" cuts. Because of this, there does not appear to be a strong relationship between price and cut. Again, it is important to note population proportions; in this attribute Astor Ideal has the lowest proportion with only 20 observations, meaning we may not be getting a clear enough picture of this data. Overall, our data does not supports Blue Nile's claim about the relationship about cut and price.

```{r, echo=FALSE}
#Box Plot of Cut vs Price
#Factoring cut
diamonds$cut <-factor(diamonds$cut, levels=c("Poor/Fair","Good","Very Good","Ideal","Astor Ideal"))

#proportion of each category
#table(diamonds$cut)

#Cut vs log(Price)
ggplot(diamonds, aes(x=cut, y=log(price)))+
  geom_boxplot()+
  theme(plot.caption = element_text(hjust = 0.5, face = "bold"))+
  labs(x="Diamond Cut", y="Log(Price)", caption="Figure 2.3.1 Box Plot of Cut Against log(Price)")
```

### 2.4 Color

#### 2.4.1 Variable Description and Claim

Third, we investigated the relationship between diamond color and price. Blue Nile claims that diamonds with less color hue are generally more highly graded and therefore more expensive. Additionally, diamonds with more of a warm color hue are generally lower rated and less expensive. In this study, diamond color is measured by grade and refers to how colorless the stone is. Grades range from the most colored to absolutely colorless is L-Z, K, J, I, H, G, F, E, D. In our data set, the color grades were fairly evenly distributed with the most common grade, F, comprising 18.4% of the data and the least common color grade , J, comprising 7.4% of the data.

```{r, echo=FALSE}
#round((table(diamonds$color) / 1214), 3)
```

#### 2.4.2 Findings

According to our bar chart in **Figures 2.4.1**, our findings show that there is a positive relationship between more colorless diamond hue and price. Diamonds that have the most colorless hue (D, E) have the highest average price compared to diamonds that have lower color grades (I, J). It is important to note that diamonds with color grade 'H' are priced higher on average than diamonds with higher colors grades (F, G). Additionally, when comparing the median prices of diamonds by color grade, the median prices between grades are relatively similar, although diamonds with higher color grades continue to have a slight price bump compared to lower diamond grades. Overall, our data somewhat supports Blue Nile's claims about the relationship between diamond color and price. 

```{r, echo=FALSE}
diamond_price_color <- diamonds %>%
  group_by(color) %>%
  summarise(mean_price = mean(price), median_price = median(price))

ggplot(data = diamond_price_color) +
  aes(x = color, y = mean_price) +
  geom_bar(stat = "identity", fill = "#16a085") +
  scale_y_continuous(labels = scales::comma) +
  theme(plot.caption = element_text(hjust = 0.5, face = "bold"))+
  labs(x = "Color", y = "Price (USD)", caption = "Figure 2.4.1. Bar graph of mean Price by Color") +
  guides(fill = "none")
```

### 2.5 Clarity

#### 2.5.1 Variable Description and Claim

Lastly, we investigated the relationship between diamond clarity and price. Blue Nile claims that diamonds with the fewest and smallest imperfections have the highest clarity grades and highest prices. Clarity refers to a diamond's imperfections either on the surface or within the stone and can be divided into 6 categories ranging from the lowest clarity grade to the highest clarity grade. In this study, diamond categories from lowest clarity grade to highest clarity grade are as follows: I3-I1, SI2-SI1, VS2-VS1, VVS2-VVS1, IF, FL. The most common clarity grade among diamonds in our sample was the 'VS1', representing 19.2% of the data. The least common clarity grade among the diamonds in our sample was the 'FL' which represented 0.2% of the data.

```{r, echo=FALSE}
#round((table(diamonds$clarity) / 1214), 3)
```

#### 2.5.2 Findings

According to the data, prices of diamonds with the highest clarity grades are not necessarily higher than other diamonds that are in lower grades. According to **Figure 2.5.1** below, diamonds with moderate clarity grades, VS1, VVS2, and VVS1, have the highest prices. High clarity diamond grades, such as IF and FL are priced lower than diamonds with moderately graded clarity. It is important to note that since the highest clarity grade, 'F', only comprises a small percentage of our data, it is lightly not accurately represented in this analysis. Based on the data that was provided, our data does not support Blue Nile's claims about the relationship between clarity and price.

```{r, echo=FALSE}
## EXPLORE CLARITY ----------
# find the number of diamonds for each clarity grade
#table(diamonds$clarity)

# create a scatterplot of price against clarity
diamonds %>% 
  ggplot(aes(x = clarity, y = price))+
  geom_point(alpha = 0.4) +
  theme(plot.caption = element_text(hjust = 0.5, face = "bold"))+
  labs(x = "Clarity", 
       y = "Price (USD)", 
       caption = "Figure 2.5.1. Scatterplot of Price against Clarity") +
  scale_x_discrete(limits = c("SI2", "SI1", "VS2", "VS1", "VVS2", "VVS1", "IF", "FL"))+
  scale_y_continuous(labels = scales::dollar)
```

### 2.6 Relationship Between Diamond Characteristics

Blue Nile makes additional claims about the relationships between each of the 4C's as well. We explore whether our data supports or contradicts those claims as well in this section.

#### 2.6.1.1. Relationship Between Cut, Carat, and Price

On one of the educational pages on the Blue Nile Website, there is a claim that diamond cut is more important that carat in regards to price. Interestingly, Blue Nile also makes the claim that carat is the most important factor that contributes to price. Since these claims are contradictory, our study investigated the relationship between cut, carat, and price. According to the scatter plot in **Figure 2.6.1.1.1**, you can see that although cut does have an effect on price, diamonds with a lower quality cut but a higher carat are generally just as expensive as diamonds with a higher quality cut and lower carat size. Our data suggests that Blue Nile's claims are unsubstantiated and that carat is a larger contributor to price than cut.

```{r, echo=FALSE}
diamonds$cut<- factor(diamonds$cut, levels = c("Astor Ideal", "Ideal", "Very Good", "Good"))

diamonds %>%
  ggplot(aes(x=log(carat), y=log(price), col=cut))+
  geom_point(alpha = 0.4)+
  theme(plot.caption = element_text(hjust = 0.5, face = "bold"))+
labs(x="log(Carat)", y="Log(Price)", caption="Figure 2.6.1.1.1. Scatterplot of log(Price) Against log(Carat)")
```

#### 2.6.1.2. Relationship Between Color, Carat, and Price

Additionally, Blue Nile claims that coloration is more visible in larger diamonds (diamonds with higher carat weight) and less noticeable in diamonds of lower carat weight. Their claim suggests that diamonds of higher carat and worse color may be valued lower than diamonds of lower carat but better color. To investigate this claim we first aimed to visualize the relationship between color and carat. According to our initial analysis, there does not appear to be a relationship between color of diamond and the carat of diamond. This data suggests that there is an even distribution of colors at each carat weight. After confirming that there is no apparent relationship between color and diamond carat, we plotted diamond price against these characteristics. According to the scatter plot in **Figure 2.6.1.2.1**, our data does not support Blue Nile’s claims about the relationship between color, carat, and price.  A diamond of lesser carat weight and worse color is still less expensive than a diamond that is of higher carat weight with the same coloration. This suggests that diamond carat weight is a greater contributor towards a diamond’s price than color.

```{r, echo=FALSE}
#initialize a new variable in the data frame
diamonds$Color_Category <- NA

diamonds$Color_Category[diamonds$color %in% c('D', 'E', 'F')] <- 1
diamonds$Color_Category[diamonds$color %in% c('G', 'H')] <- 2
diamonds$Color_Category[diamonds$color %in% c('I', 'J')] <- 3

#convert to a factor
diamonds$Color_Category <- factor(diamonds$Color_Category, levels = 1:3, labels = c("Good Color Grade (D, E, F)", "Okay Color Grade (G, H)", "Poor Color Grade (I, J)"))

diamonds %>%
  ggplot(aes(x=log(carat), y=log(price), col=Color_Category))+
  geom_point(alpha = .4)+
  theme(plot.caption = element_text(hjust = 0.5, face = "bold"))+
  labs(x="Log(Carat)", y="Log(Price)", caption="Figure 2.6.1.2.1. Scatterplot of log(Price) Against log(Carat)")
```

#### 2.6.1.3. Relationship Between Color and Clarity

Finally, Blue Nile claims that attributes such as color and clarity are somewhat related, stating "Color and relief are essentially a measure of how well a characteristic is seen. It would stand to reason that higher color grades with less tint and hues do not cover subtle imperfections that affect clarity grade as well as lower color grades do. Meaning, higher color grade diamonds should on average have lower clarity grades as more imperfections should be able to be seen. Looking at a bar chart that measures clarity frequencies against different color hues in **Figure 2.6.1.3.1.**, this does not seem to be the case however. "Colorless" hues such as E, D, and F do not seem to have a higher proportion of lower clarity grade diamonds as expected. In fact, according to **Figure 2.6.1.3.2.**, all three color categories have almost the same proportion of observations off the lowest clarity grade. This suggests that diamond color is not a contributor towards a diamond’s clarity grading, contradicting the claim made by Blue Nile.

```{r, echo=FALSE}
ggplot(diamonds, aes(x=color, fill=clarity))+
  geom_bar()+
  theme(plot.caption = element_text(hjust = 0.5, face = "bold"))+
  labs(x="Color", y="Clarity", caption="Figure 2.6.1.3.1. Boxplot of Color with Clarity frequency")
```

## 3. Simple Linear Regression

### 3.1 Overview
To fit an appropriate simple linear regression for price against carat, we first graphed a scatterplot of diamond price against carat, as shown in **Figure 3.1.1**.

```{r, echo=FALSE}
## FIT SLR MODEL ----------
# plot variables with regression line
ggplot(data = diamonds) +
  aes(x = carat, y = price) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(labels = scales::dollar) +
  theme(plot.caption = element_text(hjust = 0.5, face = "bold"))+
  labs(x = "Carat", y = "Price (USD)", caption="Figure 3.1.1. Scatter Plot of Diamond Price (USD) Against Carat")
```

To have a better proof and test if all assumptions in linear regression are met, we graphed a residual plot for the linear model of price regressed against carat, as shown in **Figure 3.1.2.1**.

```{r, included=FALSE, echo=FALSE}
# create linear model
diamond_result <- lm(price ~ carat, data = diamonds)
#summary(diamond_result)
```
```{r, echo=FALSE}
# plot residuals
par(mfrow = c(2,2))
plot(diamond_result, which=c(1,2,3,5), caption=list("Figure 3.1.2.1. Residuals vs Fitted", "Figure 3.1.2.3. Normal Q-Q", "Figure 3.1.2.2. Scale-Location", NULL, "Figure 3.1.2.4. Residuals vs Leverage"))
```

Based on the plots in **Figure 3.1.2.1.** and **Figure 3.1.2.2** and our knowledge about the data set, we found that the errors are independent and are normally distributed. However, the graphs from Figure **3.1.2.1.** and **Figure 3.1.2.2.** indicate that other two assumptions are not met. The errors do not have mean 0 (assumption 1) or have a constant variance (assumption 2).

We decided to transform the response variable (price) first to stabilize the variance. We first created a box cox plot to help us determine a range for the ideal lambda (λ) based on a 95% confidence interval. Within that confidence interval range on the box cox plot in **Figure 3.1.3.**, we chose λ = 0.3 and applied it to transform the response variable.


```{r, echo=FALSE}
# assumptions 1 and 2 are not met, so we will first transform price and then carat if neccessary

# create box cox plot
MASS::boxcox(diamond_result, 
             lambda = (seq(0, 0.5, 0.05)),
             xlab = "λ
Figure 3.1.3. Box Cox Plot of Residuals based on Figure 3.1.2.")

# we will proceed with lamba = 0.3
```

Then, we created another scatter plot of transformed price against carat in Figure **3.1.4.**, but assumption 2 was still not met. 

```{r, echo=FALSE}
# transform price
ystar_price <- (diamonds$price^0.3)
diamonds <- data.frame(diamonds, ystar_price)

ggplot(data = diamonds) +
  aes(x = carat, y = ystar_price) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(labels = scales::dollar) +
  theme(plot.caption = element_text(hjust = 0.5, face = "bold"))+
  labs(x = "Carat", y = "Transformed Price (USD)", caption="Figure 3.1.4. Scatter Plot of Transformed Price against Carat")

# assumption 2 seems to still not be met so we will re-transform price with a log transformation for variable interpretation purposes
```

Hence, we decided to re-transform the response variable with a different value of lambda. Based on **Figure 3.1.2.1.**, the errors variance was increasing as carat increased, so we needed to assign the lambda value that is less than 1. Since λ = 0.3 is close to λ = 0, we applied a log transformation on the response variable so that the results from the transformation are more easily interpreted. With both scatter plot of transformed price against carat and residual plot in **Figure 3.1.5.** and **Figure 3.1.6.1.** respectively, it was found that the log transformation provided a better fit and an easily interpretable result. However, we found that assumption 1 was still not met after solving assumption 2.

```{r, echo=FALSE}
# transform price with log
ystar_price_log <- log(diamonds$price)
diamonds <- data.frame(diamonds, ystar_price_log)

ggplot(data = diamonds) +
  aes(x = carat, y = ystar_price_log) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(labels = scales::dollar) +
  theme(plot.caption = element_text(hjust = 0.5, face = "bold"))+
  labs(x = "Carat", y = "log(Price (USD))", caption = "Figure 3.1.5. Scatter Plot of Transformed Price Against Carat")

# assumption 2 seems to be met so we will proceed with transforming carat with a log transformation for interpretation purposes
```
```{r, included=FALSE, echo=FALSE}
# transform carat
xstar_carat_log <- log(diamonds$carat)
diamonds <- data.frame(diamonds, xstar_carat_log)

# create linear model for ystar_price_log ~ xstar_carat_log
diamond_final_result <- lm(ystar_price_log ~ xstar_carat_log, data = diamonds)
#summary(diamond_final_result)

# Our R^2 value of 0.9547 indicates that 95.5% of the variability in log(price) can be accounted for by the variability in log(carat)
```

```{r, echo=FALSE}
# create linear model
diamond_result1 <- lm(ystar_price ~ carat, data = diamonds)
#summary(diamond_result1)

# plot residuals
par(mfrow = c(2,2))
plot(diamond_result1, which=c(1,2,3,5), caption=list("Figure 3.1.6.1. Residuals vs Fitted", "Figure 3.1.6.3. Normal Q-Q", "Figure 3.1.6.2. Scale-Location", NULL, "Figure 3.1.6.4. Residuals vs Leverage"))
```

Next, we transformed the predictor variable (carat) by using another log transformation based on the line shape in **Figure 3.1.5.**. We created another scatter plot with both transformed variables, as shown in **Figure 3.1.7.**, and we found that both assumptions were met after checking the residual plot in **Figure 3.1.8.1**. 

```{r, echo=FALSE}

ggplot(data = diamonds) +
  aes(x = xstar_carat_log, y = ystar_price_log) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE) +
  scale_y_continuous(labels = scales::dollar) +
  theme(plot.caption = element_text(hjust = 0.5, face = "bold"))+
  labs(x = "log(Carat)", y = "log(Price (USD))", caption = "Figure 3.1.7. Scatter Plot of Transformed Price Against Transformed Carat")
```

```{r, included=FALSE, echo=FALSE}
# create linear model for ystar_price_log ~ xstar_carat_log
diamond_final_result <- lm(ystar_price_log ~ xstar_carat_log, data = diamonds)
#summary(diamond_final_result)

# Our R^2 value of 0.9547 indicates that 95.5% of the variability in log(price) can be accounted for by the variability in log(carat)
```

```{r, echo=FALSE}
# plot residuals
par(mfrow = c(2,2))
plot(diamond_final_result, which=c(1,2,3,5), caption=list("Figure 3.1.8.1. Residuals vs Fitted", "Figure 3.1.8.3. Normal Q-Q", "Figure 3.1.8.2. Scale-Location", NULL, "Figure 3.1.8.4. Residuals vs Leverage"))
```

```{r, echo=FALSE}
# produce regression equation
#diamond_final_result

# ystar_price_log = 8.521 +1.944*xstar_carat_log

#exp(1.944)

# The price of diamonds is multiplied by $6.99 for each additional carat in the diamond
```

After all assumptions were met, we fit a new linear regression model with $y^{*}$ =`ystar_price_log` and $x^{*}$ =`xstar_carat_log` transformed variables. The final regression equation was found as:

$y^{*} = 8.521 + 1.944x^{*}$

Based on the slope of the regression equation, we interpret that the price of diamonds is multiplied by $6.99 for each additional carat in the diamond. Additionally, we found that 95.5% of the variability in log(price) can be accounted for by the variability in log(carat).

## 4. Conclusion

As we investigated the claims on the Blue Nile website regarding how diamond price is impacted by the “Four Cs”, it became clear that, while there is some element of truth to their assertion that Cut, Carat, Color, and Clarity all impact their pricing of diamonds, Carat is by far the most important variable. A diamond’s carat has an extremely strong relationship with its price and seems to be a better predictor than any of the other variables on their own or combined. For diamond shoppers looking for a large carat diamond, there is no way to avoid a large expense. For those shoppers who are flexible with their diamond carat size, though, there will be flexibility to purchase a diamond with exceptional color, clarity, and/or cut at a somewhat affordable price.





















